

\section{Additional Information}
\label{sec:duration}

\noindent \textbf{Prerequisite Knowledge} 
\label{sec:audience}
The attendees must have solid knowledge of SQL-92, since it is the baseline upon which the semistructured aspects are then added. Also solid knowledge of relational algebra. Knowledge of SQL-2003 and/or XPath are a plus but not required.

\smallskip
\noindent \textbf{Prior Related Tutorials}
In an ICDE 2015 tutorial, Ioana Manolescu and Yannis Papakonstantinou had presented algebras for the implementation of XQuery. In a SIGMOD 2016 tutorial, Yannis Papakonstantinou had presented a survey of SQL-on-Hadoop, NewSQL and NoSQL systems that support nested data query languages. The present tutorial proposal follows the methodology of the SIGMOD 2016 tutorial: The material is presented as an extension of SQL, since SQL is what is well understood by the community. The SIGMOD 2016 tutorial presented SQL extensions for complex types (nested arrays, nested tuples, etc) and explained the target languages in terms of these extensions. The present tutorial presents SQL extensions for graph data and explains the graph query languages in terms of these extensions. The present tutorial also covers some of the nested data databases covered in SIGMOD 2016 but under a different light: Instead of presenting them as complex type extensions over SQL, we present them as restrictions (into trees) of the general graph models. This allows building insight into the differences and similarities of nested and graph data models and query languages.

\smallskip
\noindent \textbf{Relevant Aspects of Presenters' Bios} 

\textbf{Alin Deutsch} is a Professor of Computer Science and Engineering at UCSD. His research pertinent to this tutorial centers around the design and optimization of semi-structured query languages
and on programming paradigms for graph data analytics. Alin is the recipient of an ACM PODS Test of Time Award, an ACM SIGMOD Top-3 Best Paper Award (together with tutorial co-author Yannis Papakonstantinou), an Alfred P. Sloan fellowship, and an NSF Career award.
He has served as Senior Scientist of TigerGraph Inc., a start-up that offers an engine capable of real-time analytics on web-scale graph data.
The analytic tasks are  expressed in a high-level graph query language called GSQL, in whose design Alin was involved.

\textbf{Yannis Papakonstantinou} is a Professor of Computer Science and Engineering at UCSD. A common theme of his research is the extension of database platforms and query processors beyond centralized relational databases and into semistructured databases, integrated views of distributed databases and web services, textual data and queries involving keyword search. His research has received more than 14,500 citations, according to Google Scholar, most of which refer to his work on semistructured data, semistructured query processing and related middleware.
In addition to his academic activity in middleware, semistructured data and query processing, Yannis was the Chief Scientist of Enosys Software, which built and commercialized an early Enterprise Information Integration platform for structured and semistructured data, utilizing XML and XQuery. The Enosys Software was OEM'd and sold under the BEA Liquid Data and BEA Aqualogic brand names. Yannis currently serves also as a consultant to Amazon Web Services.

